A Flux Scale for Southern Hemisphere 21cm EoR Experiments

### 2013-05-02 ms.tex
Edits about the beam gain with an incorrect conclusion about the error.

\begin{abstract}
We present a first catalog of broad-band spectral measurements from the 
Donald C. Backer Precision Array for Probing
% XXX ARP: was irina "first"?  and maybe "catalog"
the Epoch of Reionization (PAPER) in South Africa observed in July and
September of 2011.  In order to reduce the impact of beam calibration, which proves to
be a difficult endeavor for transit telescopes such as PAPER, on the determination
of source spectra, we have focused on calibrating sources in a narrow declination range.
Since each source follows a nearly identical path through the primary beam, this
restriction allows beam calibration to be nearly eliminated as a source of error,
yielding a dramatic improvement in the accuracy of source spectra measured in the 
100--200-MHz band that is receiving renewed attention by experiments seeking to
measure 21cm emission from the Epoch of Reionization (EoR).
Based on the variation in our data and the estimated error in the absolute
flux scale we bootstrap from multiple bright sources, we estimate PAPER measurements
of sources above XXX Jy to be accurate to 3.2\%.
When combined with catalog data at other frequencies, these data constrain parameters of
a power-law model for flux density with an uncertainty of 2.4\%.
% XXX ARP: this must again be dependent on source strength and location.
This accuracy is limited by the uncertainty in the catalog measurements we use
to estimate an absolute flux scale, and represents 
an order of magnitude improvement over previous measurements in this band.  
Comparing with prior measurements, 45 out
of 58 sources observed are found to confirm and refine a power-law model for flux density.
This includes Pictor A, which provides a key flux reference for PAPER's EoR
power spectrum analysis.
\end{abstract}

@@

One significant complication to improving the state of affairs in foreground characterization
is that many 21cm EoR experiments, including LOFAR in the northern hemisphere, and
PAPER and MWA in the southern hemisphere, are designed for drift-scan observations.
This design decision has largely been driven by the simplicity and cost-effectiveness of
phased and/or correlated dipoles for achieving the aggressive sensitivity requirements
for measuring the 21cm power spectrum of reionization (XXX cite).
Adding to the challenge, these telescopes 
cover much wider fields of view ($>10$\arcdeg) and bandwidths
($\sim 100$\% fractional) than traditional dish telescopes.
Because they do not physically point, flux calibration for such arrays relies heavily
on an accurate model of the primary beam response to correct for an
the apparent flux scale that varies across the sky.
This direction dependent gain is currently uncertain to 20\% or higher
% XXX 20% seems a bit grim.  Doesn't jonnie argue for 5 to 10% in his paper?
\citep{Jacobs:2013p9713}, approximately matching the current level of uncertainty in catalogs.

@@

Historically, the best
EoR band data were by \citet{Slee:1995p7541} with Culgoora Circular
Array\footnote{Known during daylight hours as Culgoora Radio Heliograph} and
various higher frequency measurements with Parkes.  These data are typically
uncertain to 10\% or higher and provide little coverage of the EoR band beyond
a single narrow-band data point.

@@

% XXX elaborate on:
%However, catalog source fluxes at 150MHz are also inaccurate to the 20\% level,
%particularly at southern declinations ($\delta<-20$\arcdeg).

% XXX remove PAPER specificity from below?
For power-spectrum measurements, PAPER uses an antenna layout
optimized to detect a small number of Fourier modes at high precision.  This ``maximum-redundancy''
array configuration trades imaging dynamic range for power spectrum sensitivity \citep{Parsons2012a}.
As such it is only able to easily distinguish the very brightest sources on the
sky for setting the overall flux scale \citep{Parsons2013b}. 
Of these, Pictor A is
one of the most useful, being bright, minimally resolved, and far from
confusing structure. 
For this source the current best EoR band measurement
is uncertain to 12\% and appears to imply spectral flattening in the EoR band
\citep{Perley:1997p9312}.
% XXX more background on pictor A

% XXX transition to the PAPER instrument
Establishing an accurate spectrum for Pictor A is of particular importance for
PAPER --- a dedicated EoR experiment 
that employs drift-scanning, dual-polarization dipole antennas 
tuned for efficient operation over a 120--170-MHz band.  PAPER is located in the South African Karoo desert
on the Square Kilometer Array
South Africa (SKA-SA) reserve, 100km north of the small town of Carnarvon.
The PAPER array has grown from 16 elements deployed in early 2009 to a
64-element imaging array in 2011 (see Figure \ref{fig:antpos}). 
Since November 2011 it has been arranged in a maximally redundant grid 
configuration to make deep power spectral integrations \citep{Parsons:2012a}.
% XXX to calibrate these observations...

@@

@@

% XXX incorporate
Given the shortcomings in the literature, and the necessity for PAPER power-spectrum analysis
to use Pictor A for flux calibration, we felt it prudent to examine Pictor A directly.
In this section, we describe ... % XXX

In order to ensure that 
errors in
the beam model did not skew our result, we chose a set of calibrator sources near in declination
to Pictor A that would follow nearly identical tracks through the primary beam.
This declination limit removes the necessity of correcting for a primary beam model and allows us to calculate more precise fluxes that provide a reference standard for future work.

To begin, we choose sources from the Molonglo
Reference Catalog \cite[MRC]{Large:1981p7798} that are within 5\arcdeg in
declination of a common calibrator (J2331-416) and have a flux extrapolated
from 408MHz greater than 10Jy assuming a power law spectral index of
-1.\footnote{Most radio sources in this band have power law spectra $S(\nu) =
\left(\frac{\nu}{\nu_0}\right)^\alpha$.  The 5\arcdeg distance between source
and calibrator was chosen as a balance between number of sources included and
the known smoothness of the beam.   The average spectral index for radio
sources in this band is $-0.8$.} This selection contains 62 sources in a narrow
stripe that passes through the majority of the southern EoR fields, much of the
galaxy and Centaurus A.
A list of these 
sources is given in Table \ref{tab:src_spec}.

For these sources, spectral data measured below XXX GHz were compiled from 
the NASA/IPAC Extragalactic Database (NED)
\footnote{\url{http://ned.ipac.caltech.edu/}}. Cross-matching radio sources
between bands is made difficult by changes in resolution, spectral curvature,
and resolved multi-scale structure.  For a small number of sources these
problems can be ameliorated by inspection of the spectra, source names,
coordinates,  and originating publications. However for a large sample of
sources this becomes unwieldily and prone to error. Thus for the larger number
of samples we make use of the Vollmer et al meta-catalog which takes pains to
match multi-wavelength observations \citep{Vollmer:2010p6422}.
% given the known source dimensions and spectral properties as well as telescope bandwidth and spatial resolutions.

To derive a source spectrum from each drift-scan source profile, we compute a weighted average using
weights derived from a model of
PAPER's primary beam response as an effective approximation of inverse-variance weighting
\citep{pober_et_al2012}, and averaging over the time series for each source.  Each output source spectrum
represents a best estimate given the limitations of the beam model and flux calibration that were established
up to this point, according to the equation:
% XXX equation for time-dependent beam amplitude times source flux for perceived flux density
To further improve these estimates, we derive an additional frequency-dependent gain parameter,
smoothed on 10-MHz scales,
that brings the spectrum of J2331-416 into closest agreement with a power-law spectrum
that represents the best fit to measurements below 1 GHz.
This additional gain factor was applied to all measured spectra, including that
of Pictor A.  The expectation is that, since all sources at similar declinations follow similar tracks 
through the PAPER beam, any errors present in the original measured spectra as a result of differences between
the simulated and actual beam response should be absent in the ratio between source spectra, and therefore
should be corrected by the applied gain factor.

>>>>

\begin{abstract}
We present a catalog of broad-band spectral measurements from the 
Donald C. Backer Precision Array for Probing
% XXX ARP: was irina "first"?  and maybe "catalog"
the Epoch of Reionization (PAPER) in South Africa observed in July and
September of 2011.  In order to reduce the impact of beam calibration, which proves to
be a difficult endeavor for transit telescopes such as PAPER, on the determination
of source spectra, we have focused on calibrating sources in a narrow declination range.
Since each source follows a nearly identical path through the primary beam, this
restriction allows beam calibration to be nearly eliminated as a source of error,
yielding a dramatic improvement in the accuracy of source spectra measured in the 
100--200-MHz band that is receiving renewed attention by experiments seeking to
measure 21cm emission from the Epoch of Reionization (EoR).
Based on the variation in our data and the estimated error in the absolute
flux scale we bootstrap from multiple bright sources, we estimate PAPER measurements
of sources above XXX Jy to be accurate to 3.2\%.
% DCJ 3.2% is just for Pictor A, and certainly not all PAPER measurements of sources above XXX Jy
When combined with catalog data at other frequencies, these data constrain parameters of
a power-law model for flux density with an uncertainty of 2.4\%.
% XXX ARP: this must again be dependent on source strength and location.
This accuracy is limited by the uncertainty in the catalog measurements we use
to estimate an absolute flux scale, and represents 
an order of magnitude improvement over previous measurements in this band.  
Comparing with prior measurements, 45 out
of 58 sources observed are found to confirm and refine a power-law model for flux density.
This includes Pictor A, which provides a key flux reference for PAPER's EoR
power spectrum analysis.
\end{abstract}

@@

One significant complication to improving the state of affairs in foreground characterization
is that many 21cm EoR experiments, including LOFAR in the northern hemisphere, and
PAPER and MWA in the southern hemisphere, are designed for drift-scan observations.
This design decision has largely been driven by the simplicity and cost-effectiveness of
phased and/or correlated dipoles for achieving the aggressive sensitivity requirements
for measuring the 21cm power spectrum of reionization (XXX cite).
Adding to the challenge, these telescopes 
cover much wider fields of view ($>10$\arcdeg) and bandwidths
($\sim 100$\% fractional) than traditional dish telescopes.
Because they do not physically point, flux calibration for such arrays relies heavily
on an accurate model of the primary beam response to correct for an
the apparent flux scale that varies across the sky.
This direction dependent gain is currently uncertain to 10\% or higher
% XXX 20% seems a bit grim.  Doesn't jonnie argue for 5 to 10% in his paper?
%XXX 20\% comes from the overall uncertainty between MWA and PAPER, part of which is also side lobes. 
%	Jonnie's fluxes are dominated by the side lobes in the pgb16 data, the fluxes have a range of errors around 15 to 20%
\citep{Pober:2012p8800}, and comprises a large fraction of the 20\% difference between current telescopes \citep{Jacobs:2013p9713}.

@@

Historically, the best
EoR band data were by \citet{Slee:1995p7541} with Culgoora Circular
Array\footnote{Known during daylight hours as Culgoora Radio Heliograph} and
various higher frequency measurements with Parkes.  These data are typically
uncertain to 20\% or higher and provide little coverage of the EoR band beyond
a single narrow-band data point.

@@

@@

% XXX elaborate on:
%However, catalog source fluxes at 150MHz are also inaccurate to the 20\% level,
%particularly at southern declinations ($\delta<-20$\arcdeg). 
% not sure where this went. -dcj

% XXX remove PAPER specificity from below?
EoR measurements in the southern sky have focused on the coldest regions where
galactic foregrounds are minimal. With the majority of possible observing time falling on 
RA=4h,Dec=-30. The brightest and least-resolved calibrator in this region is Pictor A 
(5h19m49.1 -45d46m45.0). Pictor A is a nearby FR-II type radio galaxy 
 similar to Cygnus A.  At $\sim$400Jy Pictor is bright and sufficiently distant from other 
bright sources to make it eminently suitable as both a phase and flux calibrator. Its apparent
size of $\sim$8\arcmin is smaller than the scales being probed by current EoR instruments , 
making resolution effects, making it suitable for precision calibration.  However, 
like most other sources, precise flux measurements in the EoR band are not available.
The current best EoR band measurement
is uncertain to 12\% and appears to imply spectral flattening in the EoR band
\citep{Perley:1997p9312}.

% XXX more background on pictor A
%  added dcj
%0518-458	 PKS, Pictor A	G	 25.93	 24.01	1		W	 0.0342	 23.63	 176.
% XXX transition to the PAPER instrument
Establishing an accurate spectrum for Pictor A is of particular importance for
PAPER --- a dedicated EoR experiment 
that employs drift-scanning, dual-polarization dipole antennas 
tuned for efficient operation over a 120--170-MHz band.  PAPER is located in the South African Karoo desert
on the Square Kilometer Array
South Africa (SKA-SA) reserve, 100km north of the small town of Carnarvon.
The PAPER array has grown from 16 elements deployed in early 2009 to a
64-element imaging array in 2011 (see Figure \ref{fig:antpos}). 
Since November 2011 it has been arranged in a maximally redundant grid 
configuration to make deep power spectral integrations \citep{Parsons:2012a}.
Though highly sensitive as a power spectrum instrument, the maximally redundant
has a broad point spread function and is able to measure only the brightest sources,
like Pictor A, for flux calibration purposes.  
% XXX to calibrate these observations...
% added dcj

@@

% XXX incorporate
Given the large uncertainties in the literature, and the necessity for PAPER power-spectrum analysis
to use Pictor A for flux calibration, we felt it prudent to examine Pictor A directly.
In this section, we describe how we use PAPER in its imaging configuration
to measure Pictor A and a selection of known, bright sources in a 5\arcdeg strip centered
on -45\arcdeg. % XXX

As we've seen above, any one point in the primary beam might be uncertain to 10\% or more. However
this is partially a function of the imaging mode. In \citet{Jacobs:2013p9837} several limitations were identified that
exasperated the primary beam uncertainty.  First, in imaging each source flux is measured at just a few 
points in the primary beam. As errors tend to vary across the beam, the uncertainty will tend to vary between sources.
Here we have averaged over all possible positions making the uncertainty a function of the net uncertainty in the beam,
and stabilizing the source-to-source variations in error bars. The second limitation identified also identified in \citet{Williams:2012p8768}
was the increase of uncertainty towards the edge of the beam. To minimize our sensitivity to both beam and noise
uncertainty, we weight the source track by an additional factor of the primary beam model.

Algebraically, the apparent flux at alittude-azimuth position $\theta-\phi$,
 is the true flux multiplied by the actual primary beam

\[
S_\textrm{apparent}(\theta,\phi) = B_A(\theta,\phi)  S
\]

Integrating over the full source track and weighting by our model primary beam, we estimate the spectrum

\begin{align}
S_\textrm{estimate} &=& \sum_t   \frac{\sum_\phi B_M(\theta,\phi) S_\textrm{apparent}(\theta,\phi) }{ \sum_\phi B^2_M(\theta,\phi)}
S_\texrm{estimate} & =&  \frac{\sum_\phi B_M(\theta,\phi) B_A(\theta,\phi)}{ \sum_\phi B^2_M(\theta,\phi)} S
\end{align}
In the fully general case, an incompletely sampled track will have its own unique weight. In this case, where each track is fully sampled,
 the only remaining independent variable is the source declination.
\[
g(\delta) = \frac{\sum_\phi B_M(\theta,\phi) B_A(\theta,\phi)}{ \sum_\phi B^2_M(\theta,\phi)}
\label{eq:g}
\]

 \begin{figure}[htbp]
\begin{center}
\includgraphics[width=0.9\textwidth]{plots/model_weighted_beam_sim_track.png}
\caption{\label{fig:beam_track_sim} Simulated model weighted beam track. }
\label{default}
\end{center}
\end{figure}

  \begin{figure}[htbp]
\begin{center}
\includgraphics[width=0.9\textwidth]{plots/model_weighted_beam_sim_track.png}
\caption{\label{fig:beam_track_sim} Simulated model weighted beam track. }
\label{default}
\end{center}
\end{figure}
 We may simulate this integral by calculating $g$ for 
similar, but not identical models. For example, consider the case where the true beam is the model beam 
but rotated by 90 degrees ('y' instead of 'x').  As shown in Figure \ref{XXX}
 tracks separated by 10\arcdeg declination differ by as much as 25\%. 
However, looking at the weighted tracks, the portion inside the sum in Eq \ref{eq:g} (Figure \ref{fig:beam_track_sim}), 
we see that  we see that over a good portion of the beam, these tracks are very similar.  Thus
  the sum displays very little declination dependance (Figure \ref{XXX}), varying by a maximum of 0.7\% over 
  the declination range -30 to -60. 
   However this conclusion is limited by the accuracy of the simulation. To limit our susceptibility to possibly larger errors we limit our use of this
  technique to with $\pm5$\arcdeg of Pictor A.

Another implication of this simulation is that there will be a net bandpass common to each source. 
We estimate this bandpass by nominating one source as a bandpass calibrator, dividing its integrated spectrum by a 
spectral model, and applying these calibration coefficients to the rest of the sources. Note that this  
only puts the measurements on a common flux scale within the PAPER band given the flux model
chosen for the calibrator.  A flux scale extrapolated from many catalog values is calculated separately in \S \ref{sec:flux_scale}.

%This limitation in declination allows us to use a single source to calibrate all the measurements without 
%needing to transfer gain solutions across the sky using a beam model. By insisting that the tracks of all sources
% overlap entirely with a common calibrator we are able to directly calibrate out the effective passband of
% the primary beam. 
%This allows us to make a more precise flux measurement that provides a reference standard for future work.

We choose sources from the Molonglo
Reference Catalog \cite[MRC]{Large:1981p7798} that are within 5\arcdeg in
declination of a common calibrator (J2331-416) and have a flux extrapolated
from 408MHz greater than 10Jy assuming a power law spectral index of
-1.\footnote{Most radio sources in this band have power law spectra $S(\nu) =
\left(\frac{\nu}{\nu_0}\right)^\alpha$. } This selection contains 62 sources in a narrow
stripe that passes through the majority of the southern EoR fields, much of the
galaxy and Centaurus A. This stripe in the broader context of the galactic plane
is shown in Figure \ref{fig:error_map}.
A list of these 
sources is given in Table \ref{tab:src_spec}.

For these sources, spectral data measured below 2 GHz were compiled from 
the NASA/IPAC Extragalactic Database (NED)
\footnote{\url{http://ned.ipac.caltech.edu/}}. Cross-matching radio sources
between bands is made difficult by changes in resolution, spectral curvature,
and resolved multi-scale structure.  For a small number of sources these
problems can be ameliorated by inspection of the spectra, source names,
coordinates,  and originating publications. However for a large sample of
sources this becomes unwieldily and prone to error. Thus for the larger number
of samples we make use of the Vollmer et al meta-catalog which takes pains to
match multi-wavelength observations \citep{Vollmer:2010p6422}.
% given the known source dimensions and spectral properties as well as telescope bandwidth and spatial resolutions.

The spectra reported here are measured by beam-forming, phasing the visibilities to
the target location, and summing to produce a dynamic spectrum. The spectrum is then isolated
from neighboring sources by filtering and averaged in the time domain.  Though we do not rely on
the primary beam model to estimate flux, we do employ it here to down weight low SNR measurements.
Each source track is paired with samples of a calibrator to build a calibrator spectrum with the same weighting
and the same track through the primary beam.  This calibrator observation provides a spectral bandpass unique 
to that particular cut through the beam.
The expectation is that, since all sources at similar declinations follow similar tracks 
through the PAPER beam, any errors present in the original measured spectra as a result of differences between
the simulated and actual beam response should be absent in the ratio between source spectra, and therefore
should be corrected by the applied gain factor.

To derive a source spectrum from each drift-scan source profile, we compute a weighted average using
weights derived from a model of
PAPER's primary beam response as an effective approximation of inverse-variance weighting
\citep{pober_et_al2012}, and averaging over the time series for each source.  Each output source spectrum
represents a best estimate given the limitations of the beam model and flux calibration that were established
up to this point, according to the equation:
% XXX equation for time-dependent beam amplitude times source flux for perceived flux density
To further improve these estimates, we derive an additional frequency-dependent gain parameter,
smoothed on 10-MHz scales,
that brings the spectrum of J2331-416 into closest agreement with a power-law spectrum
that represents the best fit to measurements below 1 GHz.
This additional gain factor was applied to all measured spectra, including that
of Pictor A.
